#Libraries
```{r}
library(stringr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(randomForest)
library(pheatmap)
library(factoextra)
library(pROC)
```

#Read in Olink and clinical data (COVID-19; PMID: 33969320)
```{r}
clinic_info <- read.delim("MGH_COVID_Clinical_Info.txt", header = F)
npx_info <- read.delim("MGH_COVID_OLINK_NPX.txt", header = F)
```

#Clean and merge data
```{r}
clinic_df <- clinic_info %>%
  mutate(id = row_number()) %>%
  separate_rows(V1, sep = ";") %>%
  group_by(id) %>%
  mutate(col_id = row_number()) %>%
  pivot_wider(names_from = col_id, values_from = V1, names_prefix = "V") %>%
  ungroup() %>%
  select(-id)
colnames(clinic_df) <- clinic_df[1,]
clinic_df <- clinic_df %>%
  slice(-1)

npx_df <- npx_info %>%
  mutate(id = row_number()) %>%
  separate_rows(V1, sep = ";") %>%
  group_by(id) %>%
  mutate(col_id = row_number()) %>%
  pivot_wider(names_from = col_id, values_from = V1, names_prefix = "V") %>%
  ungroup() %>%
  select(-id)
colnames(npx_df) <- npx_df[1,]
npx_df <- npx_df %>%
  slice(-1)

subject_npx <- npx_df %>% filter(subject_id != "") %>% 
  filter(QC_Warning != "WARN") %>%
  filter(Assay_Warning != "WARN")
clinical_npx <- merge(subject_npx, clinic_df, by = "subject_id")
```

#Identify proteins that are significant in patients with severe disease outcomes on Day 0
```{r}
#First identify proteins that are enriched in COVID positive patients (t-test)
clinical_day0 <- clinical_npx %>% 
  filter(Timepoint == "D0") %>%
  mutate(NPX = as.numeric(NPX))
proteins <- unique(clinical_day0$Assay)
COVID_biomarker <- data.frame(Protein = character(),
                      p_value_COVID = numeric(),
                      stringsAsFactors = FALSE)
for (i in proteins) {
  data_subset <- clinical_day0 %>% 
    filter(Assay == i)
  positive <- data_subset %>% filter(COVID == "1")
  negative <- data_subset %>% filter(COVID == "0")
  
  ttest <-t.test(as.numeric(positive$NPX), as.numeric(negative$NPX))
  
  COVID_biomarker <- rbind(COVID_biomarker, data.frame(
   Assay = i,
   p_value_NPX = ttest$p.value))
}
COVID_biomarker <- COVID_biomarker %>%
  mutate(adj_p_NPX = p.adjust(p_value_NPX, method = "BH")) %>%
  filter(adj_p_NPX < 0.05)

#Identify proteins that may mark severe disease outcome (linear regression)
COVID_positive <- clinical_day0 %>% 
  filter(COVID == "1") %>%                                  
  select(subject_id, Assay, NPX, Acuity_max) %>%             
  group_by(Assay, Acuity_max, subject_id) %>%                
  summarise(mean_value = mean(NPX, na.rm = TRUE)) %>%        
  ungroup()
acuity_list <- list()
for (i in proteins) {
  df_subset <- COVID_positive %>% 
    filter(Assay == i) %>%                                  
    group_by(Assay)                                          
  Acuity_model <- df_subset %>%
    summarize(p_value = summary(lm(as.integer(Acuity_max) ~ as.numeric(mean_value), data = .))$coefficients[2, 4])
  acuity_list[[i]] <- Acuity_model
}
acuity_results_df <- bind_rows(acuity_list, .id = "Assay") %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"))

Acuity_biomarkers <- acuity_results_df %>% 
  filter(p_adj < 0.05)

#Intersect proteins that are enriched in disease state and display significance in acuity
Biomarkers_of_interest <- merge(Acuity_biomarkers, COVID_biomarker, by = "Assay")
top5 <- Biomarkers_of_interest %>%
  arrange(desc(p_adj)) %>% 
  slice_tail(n = 5)       
top5_plotting <- clinical_day0 %>%
  filter(Assay %in% top5$Assay) %>% 
  select(subject_id, Assay, NPX, Acuity_max, COVID) %>%
  filter(COVID == "1") %>%
  group_by(subject_id, Assay, Acuity_max, COVID) %>%
  summarise(mean_value = mean(NPX, na.rm = TRUE))
  
#Make a boxplot of the 5 most signficant Proteins  
ggplot(top5_plotting, aes(x = Acuity_max, y = mean_value, color = Assay)) +
  geom_boxplot() +
  xlab(expression("Maximum Acuity")) +
  ylab(expression("NPX")) +
  theme(axis.text = element_text(size = 14), axis.title = element_text(size = 20), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) 
```

#Make a heatmap of significant proteins
```{r}
Pheatmap_biomarkers <- clinical_day0 %>% 
  filter(Assay %in% Biomarkers_of_interest$Assay) %>%
  filter(COVID == "1") %>%
  group_by(Assay, Acuity_max) %>% 
  summarise(mean_value = mean(as.numeric(NPX)))
Pheatmap_biomarkers_matric <- Pheatmap_biomarkers %>% 
  pivot_wider(names_from = Acuity_max, values_from = mean_value)
rownames_Pheatmap_biomarkers_matric <- Pheatmap_biomarkers_matric$Assay
Pheatmap_biomarkers_matric$Assay <- NULL
rownames(Pheatmap_biomarkers_matric) <- rownames_Pheatmap_biomarkers_matric

#Find optimal number of clusters and cluster the biomarkers by Max acuity
fviz_nbclust(Pheatmap_biomarkers_matric, kmeans, method = "wss")
set.seed(100221)
clusters <- pheatmap(Pheatmap_biomarkers_matric, cluster_cols = FALSE, kmeans_k = 4)
clusters_info <- clusters$kmeans$cluster
Pheatmap_biomarkers_matric_cluster <- cbind(Pheatmap_biomarkers_matric, clusters_info)
cluster3 <- cbind(Pheatmap_biomarkers_matric, clusters_info) %>%
  filter(clusters_info == "3")
cluster_3_fullInfo <- clinical_npx

#Add kmean clusters to the heatmap
cluster_assignment <- as.factor(Pheatmap_biomarkers_matric_cluster$clusters_info)
row_annotation <- data.frame(Cluster = cluster_assignment)
copy_matrix <- Pheatmap_biomarkers_matric
rownames(row_annotation) <- rownames(copy_matrix)
final_pheat <- pheatmap(copy_matrix, cluster_cols = FALSE, show_rownames = FALSE, annotation_row = row_annotation, annotation_colors = list(Cluster = c("1" = "cadetblue2", "2" = "darkorchid1", "3" = "red2", "4" = "darkolivegreen2")))
```

#Make a model with all biomarkers
#Make a model to predict of the 469 biomarkers play the most important role in prediciton. 
```{r}
#Generate a dataframe of the 469 proteins previously determined to be COVID biomarkers and associated with acuity
##Remove missing data
full_model <- clinical_day0 %>%
  filter(Assay %in% Biomarkers_of_interest$Assay) %>%
  filter(COVID == "1") %>%
  select(subject_id, Assay, NPX, Acuity_max) %>%
  group_by(subject_id, Assay, Acuity_max) %>% 
  summarise(mean_value = mean(as.numeric(NPX))) %>%
  ungroup %>%
  mutate(Acuity_max = ifelse(Acuity_max %in% c("1", "2"), 1,
                             ifelse(Acuity_max %in% c("3", "4", "5"), 0, NA))) %>%
  pivot_wider(names_from = Assay, values_from = mean_value) %>%
  select(-1) %>%
  na.omit

#Divide data into training and test data
set.seed(100221)
training_indicies <- sample(c(1,0), size = nrow(full_model), replace = TRUE, prob = c(0.7, 0.3))
set.seed(100221)
rf_data_full <- as.data.frame(cbind(full_model, training_indicies))
rf_data_full_train <- rf_data_full %>%
  filter(training_indicies == "1") %>%
  select(-training_indicies)
rf_data_full_test <- rf_data_full %>%
  filter(training_indicies == "0") %>%
  select(-training_indicies)
```

#Optomize random forest parameters
```{r}
ntree_values <- c(seq(from = 100, to = 500, by = 20))
nodesize_values <- c(seq(from = 2, to = 20, by = 2))
maxnode_values <- c(seq(from = 2, to = 20, by = 2))

results <- data.frame(ntree = integer(), nodesize = integer(), maxnodes = integer(), accuracy_train = numeric(), accuracy_test = numeric())

set.seed(100221)
for (ntree in ntree_values) {
  for (nodesize in nodesize_values) {
    for (maxnodes in maxnode_values) {
      
      # Initialize the accuracy vectors for each parameter combination
      acc_new <- c()
      acc_train <- c()
      
      for (i in 1:10) {
        # Train Random Forest model
        rf_model_full <- randomForest(as.factor(Acuity_max) ~ ., data = rf_data_full_train, importance = TRUE, ntree = ntree, nodesize = nodesize, maxnodes = maxnodes)
        
        # Make predictions
        predictions_train <- predict(rf_model_full, rf_data_full_train)
        predictions_test <- predict(rf_model_full, rf_data_full_test)
        
        # Calculate accuracy for training and test sets
        accuracy_train_full <- sum(predictions_train == rf_data_full_train$Acuity_max) / nrow(rf_data_full_train)
        accuracy_new_full <- sum(predictions_test == rf_data_full_test$Acuity_max) / nrow(rf_data_full_test)
        
        # Store the accuracy values for each iteration
        acc_new[i] <- accuracy_new_full
        acc_train[i] <- accuracy_train_full
      }
      
      # Compute the mean accuracy for this parameter combination
      acc_train_mean <- mean(acc_train)
      acc_new_mean <- mean(acc_new)
      
      # Append the results for this parameter combination to the results data frame
      results <- rbind(results, data.frame(ntree = ntree, nodesize = nodesize, maxnodes = maxnodes, accuracy_train = acc_train_mean, accuracy_test = acc_new_mean))
    }
  }
}

```

#Generate using a random forest model using all data points using the maximum testing accuracy parameters
```{r}
set.seed(100221)
acc_new <- c()
acc_train <- c()
meanDecreaseAcc <- matrix(NA, nrow = 100, ncol = ncol(rf_data_full_train) - 1)
meanDecreaseGini <- matrix(NA, nrow = 100, ncol = ncol(rf_data_full_train) - 1)
proteinMatrix <- matrix(NA, nrow = 100, ncol = ncol(rf_data_full_train) - 1)
for (i in 1:100) {
rf_model_full <- randomForest(as.factor(Acuity_max) ~ ., data = rf_data_full_train, importance = TRUE, ntree = 440, nodesize = 6, maxnodes = 14)
importance <- importance(rf_model_full)
Proteins <- rownames(importance)
importance <- cbind(importance, Proteins)

#Visualize the importance 
varImpPlot(rf_model_full, n.var = 10)

#Test on training data
predictions_train_full <- predict(rf_model_full, rf_data_full_train)
confusion_matrix_train_full <- table(predictions_train_full, rf_data_full_train$Acuity_max)
accuracy_train_full <- sum(diag(confusion_matrix_train_full))/sum(confusion_matrix_train_full)

#Test on new data
predictions_new_full <- predict(rf_model_full, rf_data_full_test)
confusion_matrix_new_full <- table(predictions_new_full, rf_data_full_test$Acuity_max)
accuracy_new_full <- sum(diag(confusion_matrix_new_full))/sum(confusion_matrix_new_full)
acc_new[i] <- accuracy_new_full
acc_train[i] <- accuracy_train_full
meanDecreaseAcc[i, ] <- importance[, "MeanDecreaseAccuracy"]
meanDecreaseGini[i, ] <- importance[, "MeanDecreaseGini"]
proteinMatrix[i, ] <- importance[, "Proteins"]
}
```

#Access machine learning performance
```{r}
set.seed(100221)
rf_model_full <- randomForest(as.factor(Acuity_max) ~ ., data = rf_data_full_train, importance = TRUE, ntree = 150, nodesize = 15, maxnodes = 20)

set.seed(100221)
predictions_new_full <- predict(rf_model_full, rf_data_full_test)
set.seed(100221)
predicted_prod <- predict(rf_model_full, rf_data_full_test, type = "prob")

#Compute evaluation metrics 
confusion_matrix_new_full <- table(predictions_new_full, rf_data_full_test$Acuity_max)
accuracy <- sum(diag(confusion_matrix_new_full))/sum(confusion_matrix_new_full)
precision <- confusion_matrix_new_full[2,2]/sum(confusion_matrix_new_full[2,])
recall <- confusion_matrix_new_full[2,2]/sum(confusion_matrix_new_full[,2])
f1_score <- 2 * (precision * recall)/(precision + recall)

#Calculate area under the curve (AUC) and plot the ROC
roc_curve <- roc(rf_data_full_test$Acuity_max, predicted_prod[,2])
auc <- auc(roc_curve)
ROC_plot <- plot(roc_curve)
```

#Graph accuracy during training and testing
```{r}
acc_new_df <- cbind(acc_new, "new")
colnames(acc_new_df) <- c("data", "accuracy")
acc_train_df <- cbind(acc_train, "train")
colnames(acc_train_df) <- c("data", "accuracy")
acc_trainNew_graphing <- rbind(acc_new_df, acc_train_df)
ggplot(acc_trainNew_graphing, aes(x = accuracy, y = as.numeric(data))) +
  geom_boxplot() +
  xlab(expression("Data Used")) +
  ylab(expression("Accuracy")) +
  theme(axis.text = element_text(size = 14), axis.title = element_text(size = 20), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  ylim(0.7, 1)
```

#Identify the top 21 proteins of importance in Gini and Accuracy
```{r}
protein_importance <- proteinMatrix[1,]
meanAcc <- c()
meanGini <- c()
for (i in 1:ncol(proteinMatrix)) {
  meanAcc[i] <- mean(as.numeric(meanDecreaseAcc[ ,i]))
  meanGini[i] <- mean(as.numeric(meanDecreaseGini[ ,i]))
}
Acc <- as.data.frame(cbind(protein_importance, meanAcc))
Gini <- as.data.frame(cbind(protein_importance, meanGini))

Acc_graphing <- Acc %>%
  arrange(desc(as.numeric(meanAcc))) %>%
  slice_head(n = 15)
Gini_graphing <- Gini %>%
  arrange(desc(as.numeric(meanGini))) %>%
  slice_head(n = 15)
colnames(meanDecreaseAcc) <- Proteins
colnames(meanDecreaseGini) <- Proteins
meanDecreaseAcc <- as.data.frame(meanDecreaseAcc)
meanDecreaseGini <- as.data.frame(meanDecreaseGini)

Gini_graphing_final <- meanDecreaseGini %>% 
  select(all_of(Gini_graphing$protein_importance)) %>%
  pivot_longer(cols = everything(), values_to = "Value")

Acc_graphing_final <- meanDecreaseAcc %>% 
  select(all_of(Acc_graphing$protein_importance)) %>%
  pivot_longer(cols = everything(), values_to = "Value")
```

#Graph Mean Decrease Accuracy and Gini  
```{r}
ggplot(Gini_graphing_final, aes(x = name, y = as.numeric(Value))) +
  geom_boxplot() +
  xlab(expression("Protein")) +
  ylab(expression("Mean Decrease Gini")) +
  theme(axis.text = element_text(size = 14), axis.title = element_text(size = 20), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black")) 

ggplot(Acc_graphing_final, aes(x = name, y = as.numeric(Value))) +
  geom_boxplot() +
  xlab(expression("Protein")) +
  ylab(expression("Mean Decrease Accuracy")) +
  theme(axis.text = element_text(size = 14), axis.title = element_text(size = 20), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"))
```

#Olink allows up to 21 biomarkers on the Flex
```{r}
top21 <- unique(c(Gini_graphing$protein_importance, Acc_graphing$protein_importance))
gini_merger <- subset(Gini, Gini$protein_importance %in% top21)
Acc_merger <- subset(Acc, Acc$protein_importance %in% top21)
merged_gini_acc <- merge(gini_merger, Acc_merger, by = "protein_importance")
```

```{r}
ggplot(merged_gini_acc, aes(x = as.numeric(meanGini), y = as.numeric(meanAcc), color = protein_importance)) +
  geom_point() +
  xlab(expression("Mean Gini")) +
  ylab(expression("Mean Accuracy")) +
  theme(axis.text = element_text(size = 14), axis.title = element_text(size = 20), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"))
```

#Make a model to predict of the top 21 biomarkers play the most important role in prediciton. 
```{r}
#Generate a dataframe of the top 21 proteins previously determined to be COVID biomarkers and associated with acuity
##Remove missing data
top21_data <- clinical_day0 %>%
  filter(Assay %in% merged_gini_acc$protein_importance) %>%
  filter(COVID == "1") %>%
  select(subject_id, Assay, NPX, Acuity_max) %>%
  group_by(subject_id, Assay, Acuity_max) %>% 
  summarise(mean_value = mean(as.numeric(NPX))) %>%
  ungroup %>%
  mutate(Acuity_max = ifelse(Acuity_max %in% c("1", "2"), 1,
                             ifelse(Acuity_max %in% c("3", "4", "5"), 0, NA))) %>%
  pivot_wider(names_from = Assay, values_from = mean_value) %>%
  select(-1) %>%
  na.omit

#Divide data into training and test data
set.seed(100221)
training_indicies <- sample(c(1,0), size = nrow(top21_data), replace = TRUE, prob = c(0.7, 0.3))
set.seed(100221)
rf_data_top <- as.data.frame(cbind(top21_data, training_indicies))
rf_data_top_train <- rf_data_top %>%
  filter(training_indicies == "1") %>%
  select(-training_indicies)
rf_data_top_test <- rf_data_top %>%
  filter(training_indicies == "0") %>%
  select(-training_indicies)
```

#Optomize random forest parameters for the top 21 biomarkers
```{r}
ntree_values <- c(seq(from = 100, to = 500, by = 20))
nodesize_values <- c(seq(from = 2, to = 20, by = 2))
maxnode_values <- c(seq(from = 2, to = 20, by = 2))

results <- data.frame(ntree = integer(), nodesize = integer(), maxnodes = integer(), accuracy_train = numeric(), accuracy_test = numeric())

set.seed(100221)
for (ntree in ntree_values) {
  for (nodesize in nodesize_values) {
    for (maxnodes in maxnode_values) {
      
      # Initialize the accuracy vectors for each parameter combination
      acc_new <- c()
      acc_train <- c()
      
      for (i in 1:10) {
        # Train Random Forest model
        rf_model_full <- randomForest(as.factor(Acuity_max) ~ ., data = rf_data_full_train, importance = TRUE, ntree = ntree, nodesize = nodesize, maxnodes = maxnodes)
        
        # Make predictions
        predictions_train <- predict(rf_model_full, rf_data_full_train)
        predictions_test <- predict(rf_model_full, rf_data_full_test)
        
        # Calculate accuracy for training and test sets
        accuracy_train_full <- sum(predictions_train == rf_data_full_train$Acuity_max) / nrow(rf_data_full_train)
        accuracy_new_full <- sum(predictions_test == rf_data_full_test$Acuity_max) / nrow(rf_data_full_test)
        
        # Store the accuracy values for each iteration
        acc_new[i] <- accuracy_new_full
        acc_train[i] <- accuracy_train_full
      }
      
      # Compute the mean accuracy for this parameter combination
      acc_train_mean <- mean(acc_train)
      acc_new_mean <- mean(acc_new)
      
      # Append the results for this parameter combination to the results data frame
      results <- rbind(results, data.frame(ntree = ntree, nodesize = nodesize, maxnodes = maxnodes, accuracy_train = acc_train_mean, accuracy_test = acc_new_mean))
    }
  }
}

```

#See if using just 21 biomarkers is enough